s---
title: "Homework Assignment 1"
author: "Nelson Foster"
date: "February 4, 2017"
output: html_document
---

Question 1: What is the difference between a package, library, and function?

Answer to Question 1: a package is a collection of data, compiled code, and functions in a defined structure, or library. A library is a directory where R packages are stored. A function is a piece of code written to carry out a specific task or tasks. Functions can either be included in the package, or user defined. 

Question 2: What are some ways you can find more information on R packages or datasets 
Answer to Question 2: To find more information on R packages in datasets, use the help(package = "") function, and in the quotation marks, type the package name. You can also run the help.search function to query the help files on that pacakge.

Question 3:
Install the ISLR, doBy and Tidyerse packages and run the ISLR, doBy dplyr libraries.

```{r}
library(ISLR)
library(doBy)
library(dplyr)
```

There is a dataset called Auto in the ISLR Library. Create a dataframe using the Auto dataset and call it auto_frame. 

```{r}
data(Auto)
auto_frame <- data.frame(Auto)
```

Run Basic functions to get to know auto_frame

```{r}
summary(auto_frame)
fix(auto_frame)
head(auto_frame)
```

Run basic summary statistics to further explore the dataset

```{r}
pastecs::stat.desc(auto_frame)
sapply (Auto, function(x) length(unique(x)))
str(auto_frame)
```

Use the doBy summaryBy function to create a new data frame called avempg_year_cyl and reorganize teh data to determine the average mileage per gallon per year and cylinder.

```{r}
library(doBy)
avempg_year_cyl <- summaryBy(mpg ~ year + cylinders, data=Auto)
avempg_year_cyl
```

Question 4:

Use the dplyr library and the dplyr cheat sheet in the lecture 2 file on BB to produce a similar table. 

```{r}
library(dplyr)
Auto %>%
  group_by(year, cylinders) %>%
  summarise(avempg = mean(mpg))
```

Question 5:

Which did you like better and why?

I liked the dplyr way better, it seems to be more efficient, although it took quite a while to figure out. 



