---
title: "Homework Assignment 4"
author: "Nelson Foster"
date: "March 31, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Information on the NFL 2008 Field Goal Data

Dataset: nfl2008_fga.csv

Source:  advancedfootballanalytics.com

Descriptition: Results of all NFL regular season field goal attempts
for the 2008 season.

Variable Names/Descriptions
GameDate
AwayTeam
HomeTeam
qtr   (quarter, 5=overtime)
min   (minutes remaining)
sec   (seconds remaining, added to minutes)
kickteam   (Team kicking field goal)
def        (Defending Team)
down 
togo       (Yards to go for 1st down)
kicker     (ID #)
ydline     (yardline of kicking team)
name       (kicker's name)
distance   (yards)
homekick   (1 if kicker at Home, 0 if Away)
kickdiff   (Kicking team lead +, or deficit -, prior to kick)
timerem    (Time remaining in seconds, negative ==> overtime)
offscore   (kicking team's score prior to kick)
defscore   (  defense  "     "     "    "   ")
season     (2008)
GOOD       (1 if Success, 0 if Miss)
Missed     (Missed, not blocked =1, 0 ow)
Blocked    (1 if Blocked, 0 ow)



Problem A
Usint the NFL FGA Data from 2008 in the Lecture 7 folder, create a model that can "accurately" (just do your best) classify missed versus good (made) field goals using logistic regression. In doing so, answer the following questions:
-Which variables contribute most to your model

-- Distance and Time Remaining contributed most to my model. 
-What is your "hit rate," using 50% as the break point?
--My hit rate was .115.
-What is your ROC/AUC?
--My AUC was .516, and my ROC/McFadden score was .159
-Holding all other variables in your model constant (or at an average), what are the chances of making a 50 yard field goal?
--There is approximately a 60% chance of making a 50 yard field goal. 
-What additional information might be helpful?
--Data on weather conditions would have been a helpful variable to assess. 

```{r}
require(knitr)
library(ISLR)
library(dplyr)
library(ROCR)
library(pROC)
library(pscl)
library(ResourceSelection)

```

```{r}
fgdata <- read.csv("nfl2008_fga.csv")
head(fgdata)
tail(fgdata)
str(fgdata)
str(GOOD)
structure(fgdata)

fgdata$down <- factor(fgdata$down)
fgdata$qtr <- factor(fgdata$qtr)
```

```{r}
names(fgdata)
```

```{r}

#creating train and test dataframes

fgtrain <- fgdata[1:831, ]
fgtest <- fgdata[832:1039, ]

```

```{r}

#initial model
fg.model <- glm(GOOD~.,family = binomial(link = "logit"),fgtrain)
summary(fg.model)

```

```{r}
#revised models to drop extranneous variables

fgmodel2 <- glm(GOOD~ydline+distance+homekick+kickdiff+qtr,family="binomial"(link = "logit"),fgtrain)
summary(fgmodel2)
#fgmodel3 <- glm(GOOD~timerem+distance, family="binomial"(link = "logit"), fgtrain)
fgmodel3 <- glm(GOOD~distance, family="binomial"(link = "logit"), fgdata)
summary(fgmodel3)

fgmodel2.output <- exp(coef(fgmodel2))
fgmodel2.output
fgmodel3.output <- exp(coef(fgmodel3))
fgmodel3.output
```

```{r}
#analyzing the results

anova(fg.model, test = "Chisq")
anova(fgmodel2, test  = "Chisq")
anova(fgmodel3, test = "Chisq")
```





```{r}


#additional predictions and evaluation of model (hit rate)
fgpred.model3 <- predict(fgmodel3,fgtest,type='response')


fgpred.model3 <- ifelse(fgpred.model3 > 0.5,1,0)
head(fgpred.model3)
fgmodel3hit <- mean(fgpred.model3==fgtest$GOOD)
fgmodel3hit

fgtest %>% group_by(GOOD) %>%
  summarise(no_rows = length(GOOD))


head(fgtest$GOOD)
fgmodel3hit.1 <- mean(fgpred.model3==as.numeric(fgtest$GOOD))
fgmodel3hit.1

```



```{r}

#setting the prediction and reviewing false postives 
fgnewpred <- prediction(fgpred.model3,fgtest$GOOD)

fgnewpred.performance <- performance(fgnewpred, measure = "tpr",x.measure = "fpr")
#then we plot these two measures
plot(fgnewpred.performance)

fgAUC <- performance(fgnewpred, measure = "auc")
fgAUC
```







```{r}
fgxx <- data.frame(GOOD="1",distance=50)

```

```{r}
#attempt at plotting the prediction

fgxxpred <- predict(fgmodel3,type='response')
fgxxpred
head(fgxxpred)
scatter.smooth(fgxxpred , xlab = "Field Goals", ylab = "Likelihood of 50 Yard Field Goal ")
fgxx <- data.frame(distance=50)
summary(fgxx)
summary(fgmodel3)
fgxxpred.1 <- predict(fgmodel3,fgxx)
fgxxpred.1

```



```{r}
#plotting the logistic regression 
fglogit <- glm(GOOD~timerem+distance, family = binomial(link = "logit"), data = fgdata)

hoslem.test(fgdata$GOOD, fitted(fglogit))
prob=predict(fglogit, type = c("response"))
fgdata$prop=prob

fg <- roc(GOOD~prob, data=fgdata)
pR2(fglogit)
plot(fg)
```

```{r}
summary(GOOD)

```












