---
title: "Homework Assignment 2"
author: "Nelson Foster"
date: "February 11, 2017"
output: html_document
---

```{r}
library(MASS)
```

Question 1: Use the summary function to get a summary of all the variables in the Pima.te dataset.

```{r}
summary(Pima.te)
```

Question 2: Get the structure of the pima dataset. 

```{r}
str(Pima.te)
```

Question 3: Get the variables from the dataset.

```{r}
names(Pima.te)
```

Question 4: For bmi and age variables, find out the following: mean, median, maximum, minimum, range, number of observations. 

BMI
```{r}
Pima <- data.frame(Pima.te)
#Mean
mean(Pima[["bmi"]])
#Median
median(Pima[["bmi"]])
#Max
max(Pima[["bmi"]])
#Min
min(Pima[["bmi"]])
#Range
range(Pima[["bmi"]])
#Number of Observations
nrow(Pima)
```

Age
```{r}
Pima <- data.frame(Pima.te)
#Mean
mean(Pima[["age"]])
#Median
median(Pima[["age"]])
#Max
max(Pima[["age"]])
#Min
min(Pima[["age"]])
#Range
range(Pima[["age"]])
#Number of Observations
nrow(Pima)
```

Question 5: How many women are in this dataset?

Answer: There are 332 women in this dataset. 

Question 6: Select the first 5 observations and first four columns/variables from the dataset.

```{r}
Pima[1:5,1:4]
```

Question 7: Select the records where bmi is grater than or equal to 50.

```{r}
sub = subset(Pima, bmi == 50, select =c("bmi"))
sub
```

Answer: There are 70 observations of BMI scores at or above 50. 


Question 8: What percentage of the women have diabetes by WHO criteria?

```{r}
library(ggplot2)
tbl <- table(Pima$type)
cbind(tbl,prop.table(tbl))
nrow(Pima)

```
Answer: According to WHO criteria, 33% of the women in the Pima study have diabetes.

Question 9: Obtain a histogram for body mass index

```{r}
hist(Pima[["bmi"]])
```

Question 10: What are the mean and median for bmi? How far apart are they?

Mean (bmi)
```{r}
mean(Pima[["bmi"]])
```
Median (bmi)
```{r}
median(Pima[["bmi"]])
```
Distance between mean and median:
```{r}
x1 <- mean(Pima$bmi)
x2 <- median(Pima$bmi)
dist(rbind(x1, x2))
```
Answer: The mean and median are 0.34 units apart. 

Question 11: Get data from "vlbw.csv" file.

```{r}
library(readr)
vlbw <- read_csv("vlbw.csv")
```

Question 12: Obtain a histogram of the length of stay, i.e. the number of days the infants stay in the neonatal intensive care unit (variable: hospstay)

```{r}
hist(vlbw$hospstay)
```

Question 13: Do you see data problems?

Answer: Yes. According to the console output, there are missing column names in the dataset. This is creating some errant data, which is skewing histogram's range (-6000 to 4000) and the count of the hospital stays.

Question 14: Draw a boxplot of the variable lowph. Visually estimate the median and quartiles from the plot.

```{r}
boxplot(vlbw$lowph)
```


Answer: Median: approx. 7.2; Q1: approx. 7.1;  Q3: 7.3.

Question 15: Use the summary function to check your answers to the previous question. is it about the same?

```{r}
summary(vlbw$lowph)
```

Answer: Yes, there are more or less the same. 

Question 16: The variable lowph contains the lowest ph in the first 4 days of life. Obtain a histogram of this variable (the variable is called lowph)

```{r}
hist(vlbw$lowph)
```
